maze-dataset: Maze Generation with Algorithmic Variety and Representational Flexibility

### 2025-05-28 docs/paper/figures/outputs-ascii-colored.tex
wip

\begin{Verbatim}[commandchars=\\\{\}]
# # # # # # # # # # #
#         \textcolor{blue}{X X X} #   #
#   # # # \textcolor{blue}{X} # \textcolor{blue}{X} #   #
#       # \textcolor{blue}{X} # \textcolor{green}{S}     #
# # # # # \textcolor{blue}{X} # # # # #
# \textcolor{blue}{X X X X X} # \textcolor{red}{E} \textcolor{blue}{X X} #
# \textcolor{blue}{X} # # #   # # # \textcolor{blue}{X} #
# \textcolor{blue}{X} #           # \textcolor{blue}{X} #
# \textcolor{blue}{X} # # # # # # # \textcolor{blue}{X} #
# \textcolor{blue}{X X X X X X X X X} #
# # # # # # # # # # #
\end{Verbatim}

>>>>

{\ttfamily
\setlength{\tabcolsep}{0.5em}
\setlength{\arraystretch}{0.8}
\begin{tabular}{ccccccccccc}
\# & \# & \# & \# & \# & \# & \# & \# & \# & \# & \# \\
\# &   &   &   &   & \textcolor{blue}{X} & \textcolor{blue}{X} & \textcolor{blue}{X} & \# &  & \# \\
\# &   & \# & \# & \# & \textcolor{blue}{X} & \# & \textcolor{blue}{X} & \# &   & \# \\
\# &   &   &   & \# & \textcolor{blue}{X} & \# & \textcolor{green}{S} &   &   & \# \\
\# & \# & \# & \# & \# & \textcolor{blue}{X} & \# & \# & \# & \# & \# \\
\# & \textcolor{blue}{X} & \textcolor{blue}{X} & \textcolor{blue}{X} & \textcolor{blue}{X} & \textcolor{blue}{X} & \# & \textcolor{red}{E} & \textcolor{blue}{X} & \textcolor{blue}{X} & \# \\
\# & \textcolor{blue}{X} & \# & \# & \# &   & \# & \# & \# & \textcolor{blue}{X} & \# \\
\# & \textcolor{blue}{X} & \# &   &   &   &   &   & \# & \textcolor{blue}{X} & \# \\
\# & \textcolor{blue}{X} & \# & \# & \# & \# & \# & \# & \# & \textcolor{blue}{X} & \# \\
\# & \textcolor{blue}{X} & \textcolor{blue}{X} & \textcolor{blue}{X} & \textcolor{blue}{X} & \textcolor{blue}{X} & \textcolor{blue}{X} & \textcolor{blue}{X} & \textcolor{blue}{X} & \textcolor{blue}{X} & \# \\
\# & \# & \# & \# & \# & \# & \# & \# & \# & \# & \#
\end{tabular}
}

### 2025-05-28 docs/paper/paper.tex
wip

Internally, mazes are
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#SolvedMaze}{\texttt{SolvedMaze}}
objects, which have path information and a tensor optimized for storing
sub-graphs of a lattice. These objects can be converted to and from
several formats to maximize their utility in different contexts.

@@

In previous work, maze tasks have been used with Recurrent Convolutional
Neural Network (RCNN) derived architectures
(\protect\hyperlink{ref-deepthinking}{Schwarzschild, Borgnia, Gupta,
Huang, et al., 2021}). To facilitate the use of our package in this
context, we replicate the format of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}) and provide the
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html\#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}}
class which returns rasterized pairs of (input, target) mazes as shown
in \autoref{fig:e2h-raster} below.

@@

Autoregressive transformer models can be quite sensitive to the exact
format of input data, and may even use delimiter tokens to perform
reasoning steps (\protect\hyperlink{ref-pfau2024dotbydot}{Pfau et al.,
2024}; \protect\hyperlink{ref-spies2024causalworldmodels}{Spies et al.,
2024}). To facilitate systematic investigation of the effects of
different representations of data on text model performance, we provide
a variety of text output formats. By passing an instance of
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization.html\#MazeTokenizerModular}{\texttt{MazeTokenizerModular}}
to
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.as_tokens}{\texttt{as\_tokens(...)}},
a maze can be converted to a text sequence. The
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization.html\#MazeTokenizerModular}{\texttt{MazeTokenizerModular}}
class contains a rich set of options with class contains 19 discrete
parameters, resulting in over 5.8 million unique possible tokenizers.

All output sequences consist of four token regions representing
different features of the maze; an example output sequence is shown in
\autoref{fig:token-regions}.

@@

\hypertarget{benchmarks}{%
\subsection{Benchmarks of Generation Speed}\label{benchmarks}}

We benchmarks for generation time across various configurations in
\autoref{tab:benchmarks} and \autoref{fig:benchmarks}. Experiments were
performed on a
\href{https://docs.github.com/en/actions/using-github-hosted-runners/using-github-hosted-runners/about-github-hosted-runners\#standard-github-hosted-runners-for-public-repositories}{standard
GitHub runner} without parallelism.

@@

\hypertarget{sec:success-rate-estimation}{%
\subsection{Success Rate Estimation}\label{sec:success-rate-estimation}}

Maze generation under certain constraints may not always be successful,
and for this we provide a way to estimate the success rate of a given
configuration, described in \autoref{fig:sre}.

@@

Using an adjacency matrix for storing mazes would be memory inefficient
by failing to exploit the highly sparse structure, while using an
adjacency list could lead to a poor lookup time. This package utilizes a
simple, efficient representation of mazes as subgraphs of a finite
lattice, which we call a
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze}{\texttt{LatticeMaze}}.

>>>>

Internally, mazes are
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#SolvedMaze}{\texttt{SolvedMaze}}
objects, which have path information and a tensor optimized for storing
sub-graphs of a lattice. These objects can be converted to and from
several formats, shown in \autoref{fig:output-fmts}, to maximize their
utility in different contexts.

In previous work, maze tasks have been used with Recurrent Convolutional
Neural Network (RCNN) derived architectures
(\protect\hyperlink{ref-deepthinking}{Schwarzschild, Borgnia, Gupta,
Huang, et al., 2021}). To facilitate the use of our package in this
context, we replicate the format of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}) and provide the
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html\#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}}
class which returns rasterized pairs of (input, target) mazes as shown
in \autoref{fig:e2h-raster}.

@@

@@

Autoregressive transformer models can be quite sensitive to the exact
format of input data, and may even use delimiter tokens to perform
reasoning steps (\protect\hyperlink{ref-pfau2024dotbydot}{Pfau et al.,
2024}; \protect\hyperlink{ref-spies2024causalworldmodels}{Spies et al.,
2024}). To facilitate systematic investigation of the effects of
different representations of data on text model performance, we provide
a variety of text output formats. By passing an instance of
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization.html\#MazeTokenizerModular}{\texttt{MazeTokenizerModular}}
to
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.as_tokens}{\texttt{as\_tokens(...)}},
a maze can be converted to a text sequence. The
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/tokenization.html\#MazeTokenizerModular}{\texttt{MazeTokenizerModular}}
class contains a rich set of options with class contains 19 discrete
parameters, resulting in over 5.8 million unique possible tokenizers.
All output sequences consist of four token regions representing
different features of the maze; an example output sequence is shown in
\autoref{fig:token-regions}.

@@

\hypertarget{benchmarks}{%
\subsection{Benchmarks}\label{benchmarks}}

We benchmarks for generation time across various configurations in
\autoref{tab:benchmarks} and \autoref{fig:benchmarks}. Experiments were
performed on a
\href{https://docs.github.com/en/actions/using-github-hosted-runners/using-github-hosted-runners/about-github-hosted-runners\#standard-github-hosted-runners-for-public-repositories}{standard
GitHub runner} without parallelism. Additionally, maze generation under
certain constraints may not always be successful, and for this we
provide a way to estimate the success rate of a given configuration,
described in \autoref{fig:sre}.

@@

@@

Using an adjacency matrix for storing mazes would be memory inefficient
by failing to exploit the highly sparse structure, while using an
adjacency list could lead to a poor lookup time. This package utilizes a
simple, efficient representation of mazes as subgraphs of a finite
lattice, detailed in \autoref{fig:maze-impl}, which we call a
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze}{\texttt{LatticeMaze}}.
